### 2015-05-15 doc/thesis/system.tex
Removing rules that only handle expression with multiple results

First, our type system type checks the expression list in the right-hand
side of the assignment.
In our example, the right-hand side of the assignment has type
$1 \times ``foo" \times \Nil{*}$.
Note that our type system includes the type $\Nil{*}$ to replace
missing values.
The rules that type check expression lists introduce the type
$\Nil{*}$ to let the right-hand side produce fewer values than
expected in the left-hand side.
Our example uses the rule \textsc{T-EXPLIST1} to type check
the right-hand side of the assignment.
The rule \textsc{T-EXPLIST1} is the rule that type checks an
expression list where all expressions can only produce a single value:
\[
\begin{array}{c}
\mylabel{T-EXPLIST1}\\
\dfrac{\env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
       n = |\;\overline{e}\;|}
      {\env_{1}, \penv \vdash \overline{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{m}}
\end{array}
\]

@@

After type checking the right-hand side, our type system type checks the list
of left-hand values.
In our example, the left-hand side of the assignment has type 
$\Integer \times \String \times \Value{*}$.
Note that our type system uses the type $\Value{*}$ to discard extra values.
The rule that type checks lists of left-hand values introduces the type
$\Value{*}$ to let the right-hand side produce more values than
expected in the left-hand side.
Our example uses the rule \textsc{T-LHSLIST} to type
check a list of left-hand values:
\[
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env_{1}, \penv \vdash l_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
       n = |\;\overline{l}\;|}
      {\env_{1}, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
\end{array}
\]

@@

Our type system includes $\Nil{*}$ in the type of an expression list only
if its type does not end in another variadic type $T{*}$.
For instance, rules \textsc{T-EXPLIST2} and \textsc{T-EXPLIST3} handle
the case where an expression list is just an expression that may produce
multiple values.
The former rule includes the type $\Nil{*}$, but the latter rule does not:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-EXPLIST2}\\
\dfrac{\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Void, \env_{2}}
      {\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-EXPLIST3}\\
\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n}{*}, \env_{1}
\end{array}
\end{array}
\]

@@

The rule \textsc{T-CONSTRUCTOR2} uses these steps to type check a
table constructor with a non-empty list of table fields:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(K_{i},V_{i}), \env_{i+1} \\
       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
       wf(T) \;\;\;
       n = |\;\overline{[e_{1}] = e_{2}}\;| \\
       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) 
       \end{array}}
      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{m}}
\end{array}
\]

@@

First, our type system uses the auxiliary relation
$\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$
for type checking $idiv(1, 2)$.
This relation means that given a type environment $\env_{1}$ and
a projection environment $\penv$, we can check that an expression
list $el$ has type $S_{1}$ and produces a new type environment $\env_{2}$
and produces a pair $(x,S_{2})$.
This pair means that the last expression of an expression list $el$
produces an union of second-level types $S_{2}$ that should be bound
to a variable $x$ in the projection environment $\penv$,
as the resulting type of this expression is a tuple of projection
types $\pi_{i}^{x}$.
In our example, our type system uses rule \textsc{T-EXPLIST6} for
type checking $idiv(1, 2)$:
\[
\begin{array}{c}
\mylabel{T-EXPLIST6}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Void \sqcup T_{1}' \times ... \times T_{n}' \times \Void, \env_{2}\\
       S = T_{1} \times ... \times T_{n} \times \Nil{*} \sqcup T_{1}' \times ... \times T_{n}' \times \Nil{*}
       \end{array}}
      {\env_{1}, \penv \vdash me:\pi_{1}^{x} \times ... \times \pi_{n}^{x} \times \Nil{*}, \env_{2}, (x,S)}
\end{array}
\]

>>>>

First, our type system type checks the expression list in the right-hand
side of the assignment.
In our example, the right-hand side of the assignment has type
$1 \times ``foo" \times \Nil{*}$.
Note that our type system includes the type $\Nil{*}$ to replace
missing values.
The rules that type check expression lists introduce the type
$\Nil{*}$ to let the right-hand side produce fewer values than
expected in the left-hand side.
Our example uses the rule \textsc{T-EXPLIST1} to type check
the right-hand side of the assignment.
The rule \textsc{T-EXPLIST1} is the rule that type checks an
expression list where all expressions can only produce a single value:
\[
\begin{array}{c}
\mylabel{T-EXPLIST1}\\
\dfrac{\env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{f} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
       n = |\;\overline{e}\;|}
      {\env_{1}, \penv \vdash \overline{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{f}}
\end{array}
\]

@@

After type checking the right-hand side, our type system type checks the list
of left-hand values.
In our example, the left-hand side of the assignment has type 
$\Integer \times \String \times \Value{*}$.
Note that our type system uses the type $\Value{*}$ to discard extra values.
The rule that type checks lists of left-hand values introduces the type
$\Value{*}$ to let the right-hand side produce more values than
expected in the left-hand side.
Our example uses the rule \textsc{T-LHSLIST} to type
check a list of left-hand values:
\[
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env_{1}, \penv \vdash l_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{f} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
       n = |\;\overline{l}\;|}
      {\env_{1}, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{f}}
\end{array}
\]

@@

Our type system includes $\Nil{*}$ in the type of an expression list only
if its type does not end in another variadic type $T{*}$.
For instance, the rules \textsc{T-EXPLIST2} and \textsc{T-EXPLIST3} handle
the case where an expression list ends with an expression that may produce
multiple values.
The former rule includes the type $\Nil{*}$, but the latter rule does not:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-EXPLIST2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{1}, \penv \vdash me:T_{n+1} \times ... \times T_{n+m} \times \Void, \env_{n+2}\\
       \env_{f} = merge(\env_{1}, ..., \env_{n+2}) \;\;\;
       n = |\;\overline{e}\;|
       \end{array}}
      {\env_{1}, \penv \vdash \overline{e},me:T_{1} \times ... \times T_{n+m} \times \Nil{*}, \env_{f}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-EXPLIST3}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{1}, \penv \vdash me:T_{n+1} \times ... \times T_{n+m}{*}, \env_{n+2}\\
       \env_{f} = merge(\env_{1}, ..., \env_{n+2}) \;\;\;
       n = |\;\overline{e}\;|
       \end{array}}
      {\env_{1}, \penv \vdash \overline{e},me:T_{1} \times ... \times T_{n+m}{*}, \env_{f}}
\end{array}
\end{array}
\]

@@

The rule \textsc{T-CONSTRUCTOR2} uses these steps to type check a
table constructor with a non-empty list of table fields:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(K_{i},V_{i}), \env_{i+1} \\
       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
       wf(T) \;\;\;
       n = |\;\overline{[e_{1}] = e_{2}}\;| \\
       \env_{f} = merge(\env_{1}, ..., \env_{n+1})
       \end{array}}
      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{f}}
\end{array}
\]

@@

First, our type system uses the auxiliary relation
$\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$
for type checking $idiv(1, 2)$.
This relation means that given a type environment $\env_{1}$ and
a projection environment $\penv$, we can check that an expression
list $el$ has type $S_{1}$ and produces a new type environment $\env_{2}$
and produces a pair $(x,S_{2})$.
This pair means that the last expression of an expression list $el$
produces an union of second-level types $S_{2}$ that should be bound
to a variable $x$ in the projection environment $\penv$,
as the resulting type of this expression is a tuple of projection
types $\pi_{i}^{x}$.
In our example, our type system uses rule \textsc{T-EXPLIST4} for
type checking $idiv(1, 2)$:
\[
\begin{array}{c}
\mylabel{T-EXPLIST4}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{1}, \penv \vdash me:S, \env_{n+2}\\
       S = T_{n+1} \times ... \times T_{n+m} \times \Void \sqcup T_{n+1}' \times ... \times T_{n+m}' \times \Void\\
       \env_{f} = merge(\env_{1}, ..., \env_{n+2}) \;\;\;
       n = |\;\overline{e}\;|
       \end{array}}
      {\env_{1}, \penv \vdash \overline{e},me:T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x} \times \Nil{*}, \env_{f}, (x,S)}
\end{array}
\]
